The Sub-Millimetre User Reduction Facility: an iterative map-maker for \scuba

### 2012-06-14 smurf_paper.tex
complete draft of intro

Submillimetre imaging cameras generally use bolometers to maximize
sensitivity, in which an absorber collects photons that pass through
broad-band filters, and the signal is read out-out using a thermometer
\citep[e.g.][]{griffin2002}. The sensitivity of bolometers in a camera
like \scuba\ are limited by photon noise from the instrument and
ambient backgrounds. The low-frequency noise, however, is dominated by
sources such as slow variations in mean backgrounds (e.g., thermal
variations within the cryostat, and the atmosphere for ground-based
cameras), and drift in the readout electronics. Such noise has a power
spectrum $\propto 1/f^\alpha$ ($\alpha>0$), and the frequency at which
it is comparable to the white noise level is called the ``$1/f$
knee''. Since the low-frequency noise is largely correlated between
all, or subsets, of the bolometers in time, it can be suppressed
during map-making, since astronomical signals have the distinct
property that they are fixed in a sky reference frame (assuming they
are not time-varying). If successful, the noise in the resulting map
is said to be ``white noise limited'', meaning that the map noise is
uncorrelated spatially, and has an amplitude that scales as the
$\mathrm{NEFD}/\sqrt{t}$, where the NEFD is the noise-equivalent
flux-density (the white noise level of a bolometer in 1\,s of
integration), and $t$ is the amount of integration time in a map
pixel.

@@

There are three general styles of map-making that are relevant to
reducing bolomter data in the literature. The simplest methods involve
some basic processing of the data to remove as much noise
contamination as possible, e.g., using baseline removal, filters, and
then simply re-gridding the data into a map. Such was the basic recipe
for the baseline reduction of chopped data from SCUBA-2's predecessor
SCUBA (reference?), and MAMBO, another camera from the same
generation. Generally speaking, such methods are fast, although
dependending on the science goals and noise properties of the data,
they may not achieve the best theoretical noise performance on the
angular scales of interest. A particular case of this method was
successfully used for the reduction of data from BOLOCAM
\citep[e.g.][]{laurent2005}, and its younger sibling AzTEC
\citep[e.g.][]{scott2008}, using ``principal component analysis''
(PCA) cleaning. A statistical black-box is used to remove the most
correlated components of the bolometer signals, enabling the detection
of point-sources very close to the theoretical white-noise limits of
the detectors with reasonable computation times. However, PCA cleaning
is not a good generic solution for producing maps of extended
structures, since large astronomical sources produce correlated
signals amongst many detectors, and are removed by this procedure.

@@

\scuba\ is presently unique in terms of the shear volume of data that
it produces on a nightly basis, and the fact that it is observing a
wide variety of science targets for multiple users (unlike dedicated
CMB experiments, for example). For this reason, SMURF has been
designed to use a faster, and less memory-intensive algorithm than
CMB-style maximum likelihood map-makers, although it is less accurate
at reproducing large-scale structures. The basic procedure is to
iteratively model and remove most of the bolometer-bolometer
correlated signal components and independent low-frequency noise, and
then regrid the residual signals. This method is similar in spirit to
the Comprehensive Reduction Utility for SHARC-2
\citep[CRUSH,][]{kovacs2008}, and the pipeline developed for the
Bolocam Galactic Plane Survey \citep{aguirre2010}. While most of the
analysis described here applies equally to all of the data taken
to-date with \scuba, we note that the \scuba\ Shared-Risks Observing
period (S2SRO) which took place from 2010 February 11 to 2010 March 25
used only a single subarray at each of the two observing wavelengths,
and had considerably different noise properties and poorer performance
than the fully-commissioned instrument in 2011.

\textit{May want to note that the pipeline is discussed elsewhere,
  e.g. Gibb et al 2006.}

@@

%In the case of chopped data, ideally the resulting data are completely
%free of low-frequency noise, and therefore the data samples are
statistically uncorrelated.

@@

%-------------------------------------------------
\subsection{Differences between the S2SRO and 2011 re-commissioned instrument}
%-------------------------------------------------

>>>>

Submillimetre imaging cameras generally use bolometers to maximize
sensitivity, in which an absorber collects photons that pass through
broad-band filters, and the signal is read out-out using a thermometer
\citep[e.g.][]{griffin2002}. The sensitivity of bolometers in a camera
like \scuba\ are limited by white photon and phonon noise from the
instrument and ambient backgrounds. The low-frequency noise, however,
is dominated by sources such as slow variations in mean backgrounds
(e.g., thermal variations within the cryostat, and the atmosphere for
ground-based cameras), and drift in the readout electronics. Such
noise has a power spectrum $\propto 1/f^\alpha$ ($\alpha>0$), and the
frequency at which it is comparable to the white noise level is called
the ``$1/f$ knee''. Since the low-frequency noise is largely
correlated between all, or subsets, of the bolometers in time, it can
be suppressed during map-making, since astronomical signals have the
distinct property that they are fixed in a sky reference frame
(assuming they are not time-varying). If successful, the noise in the
resulting map is said to be ``white noise limited'', meaning that the
map noise is uncorrelated spatially, and has an amplitude that scales
as the $\mathrm{NEFD}/\sqrt{t}$, where the NEFD is the
noise-equivalent flux-density (the white noise level of a bolometer in
1\,s of integration), and $t$ is the amount of integration time in a
map pixel.

@@

There are three general styles of map-making that are relevant to
reducing bolomter data in the literature. The simplest ``direct
methods'' involve some basic processing of the data to remove as much
noise contamination as possible, e.g., using baseline removal,
filters, and then simply re-gridding the data into a map. Such was the
basic recipe for the baseline reduction of chopped data from SCUBA-2's
predecessor SCUBA (reference?), and MAMBO, another camera from the
same generation. Generally speaking, such methods are fast, although
dependending on the science goals and noise properties of the data,
they may not achieve the best theoretical noise performance on the
angular scales of interest. A particular case of this method was
successfully used for the reduction of data from BOLOCAM
\citep[e.g.][]{laurent2005}, and its younger sibling AzTEC
\citep[e.g.][]{scott2008}, using ``principal component analysis''
(PCA) cleaning. A statistical black-box is used to remove the most
correlated components of the bolometer signals, enabling the detection
of point-sources very close to the theoretical white-noise limits of
the detectors with reasonable computation times. However, PCA cleaning
is not a good generic solution for producing maps of extended
structures, since large astronomical sources produce correlated
signals amongst many detectors, and are removed by this procedure.

@@

The third approach adopted here for \scuba\ is a compromise between
the previous two methods. Under the assumption that a significant
portion of the (predominantly low-frequency) non-white noise sources
can be modeled, an iterative solution is obtained for both the
astronomical image and the parameters of the noise model. Since the
remaining (non-modelled) noise sources are assumed to be white, a
single scalar noise may be calculated for all of the data points from
a given bolometer (since the noise at any instant in time is
uncorrelated with others, nor with the data from other bolometers),
greatly simplifying the measurement of noise properties and the
inversion step that is so complicated in the maximum-likelihood
methods. In principle, it should also be possible to recover a wider
dynamic range in angular scales than the simplest ``direct'' methods,
with an increase in calculation time that is merely linear in the
number of iterations needed, while using a comparable amount of
memory. A fast iterative method for handling chopped data from CMB
experiments was described in \citet{wright1996}, and later applied to
chopped SCUBA data by \citet{johnstone2000}, although in each case the
chopping renders the data essentially white without the need to model
additional noise components. More similar to the algorithm described
here was the iterative method for fitting and removing baseline drifts
in SCUBA scan-map data as the astronomical image estimate improved
(...SURF reference?...). The two closest modern relatives of SMURF are the
Comprehensive Reduction Utility for SHARC-2
\citep[CRUSH,][]{kovacs2008}, and the pipeline developed for the
Bolocam Galactic Plane Survey \citep{aguirre2010}.

A reasonable model for correlated noise in \scuba\ is a single
``common-mode'' signal seen by all of the bolometers. Iterative
estimation and removal of this signal significantly lowers the $1/f$
knee, without compromosing structures on angular scales larger than
the array footprint. Residual independent drifts at lower frequencies
are removed with an iterative high-pass filter. This strategy enables
SMURF to reduce data faster than they are taken, on single high-end
desktop computers, and it has successfully been used as part of a
real-time pipeline offering feedback to observers at the
telescope. The pipeline is also used to generate near-science grade
products at the Canadian Astronomical Data Centre (CADC). Details of
the pipeline design are given in \citet{gibb2005}

This paper is divided in the following way. We first describe the
properties of \scuba\ data and scan patterns in Section~\ref{sec:data}
to motivate the data model that we have adopted. Note that while most
of the analysis described here applies equally to all of the data
taken to-date with \scuba, the \scuba\ Shared-Risks Observing period
(S2SRO) which took place from 2010 February 11 to 2010 March 25 used
only a single subarray at each of the two observing wavelengths, and
had considerably different noise properties and poorer performance
than the fully-commissioned instrument in 2011 (a comparison is
provided in Section~\ref{sec:s2sro}. Next, the details of the SMURF
algorithm are given in Section~\ref{sec:algorithm}. The paper is
concluded in Section~\ref{sec:examples} with three detailed test cases
that span the majority of observation types likely to be undertaken
with \scuba, with an emphasis on the mitigation of divergence
problems, and characterizing the output maps: (i) Uranus, a bright,
compact source (Section~\ref{sec:point}); (ii) the Lockman Hole, a
blind survey of faint point-like source (Section~\ref{sec:cosmo}); and
(iii) the star-forming region M17, including bright, extended emission
(Section~\ref{sec:extended}). All of the data analyzed in this paper
are publicly available through the CADC SCUBA-2 raw-data queries page
for the dates and observation numbers given in the
text\footnote{\texttt{http://www3.cadc-ccda.hia-iha.nrc-cnrc.gc.ca/
jcmt/search/scuba2}}. All of the analysis was performed using
\texttt{Starlink} version ... .

%\scuba\ is presently unique in terms of the shear volume of data that
%it produces on a nightly basis, and the fact that it is observing a
%wide variety of science targets for multiple users (unlike dedicated
%CMB experiments, for example). For this reason, SMURF has been
%designed to use a faster, and less memory-intensive algorithm than
%CMB-style maximum likelihood map-makers, although it is less accurate
%at reproducing large-scale structures. The basic procedure is to
%iteratively model and remove most of the bolometer-bolometer
%correlated signal components and independent low-frequency noise, and
%then regrid the residual signals.

@@

%In the case of chopped data, ideally the resulting data are completely
%free of low-frequency noise, and therefore the data samples are
%statistically uncorrelated.

@@

%-------------------------------------------------
\subsection{Differences between the S2SRO and 2011 re-commissioned instrument}
\label{sec:s2sro}
%-------------------------------------------------
